\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-05-29 submission/manuscript.Rmd
Review optimization.

Chloroplast sequence abundances were higher in autumn/winter than in spring/summer. This pattern is not surprising as seagrasses harbor more algal epiphytes during autumn/winter than in spring/summer [@Reyes2001]. Furthermore, we used an adapted DNA isolation protocol that is known to partially co-extract DNA from planktonic eukaryotes [@Korlevic2015]. Strong seasonal fluctuations of high rank epiphytic taxa were not observed, with the exception of *Cyanobacteria*. Cyanobacterial sequences were more pronounced in November and December than in spring and summer. In the months of high cyanobacterial sequence abundances the majority of sequences from this group were classified as *Pleurocapsa*, a group known to colonized different living and non-living surfaces [@Longford2007; @Burns2004; @Mobberley2012; @Reisser2014]. It is possible that during periods of low metabolic activity there is a reduced interaction and selection of the epiphytic community by the seagrass, causing leaves to become a suitable surface for nonspecific colonizers [@Zavodnik1998]. *Pleurocapsa* was replaced in spring and summer by *Acrophormium*, *Phormidesmis* and sequences without known relatives within the *Nodosilineaceae*. A study of coastal microbial mats found also a higher proportion of *Nodosilineaceae* sequences in summer, while *Phormidesmis* sequences were at their peak in autumn [@Cardoso2019]. Other high rank taxa did not exhibit strong successional patterns. In every analyzed group, with the exception of *Desulfobacterota*, taxa present throughout the year in similar proportions and season specific taxa could be identified (Figs. \ref{bactero} and \ref{desulfo}). Within *Bacteroidota* different groups within the family *Saprospiraceae* (e.g. *Lewinella*, *Portibacter* and *Rubidimonas*) were detected across all seasons. Members of this family are often found in association with macrophytes and it is suggested that they are involved in the hydrolysis and utilization of complex carbon sources [@Burke2011a; @Crump2018; @McIlroy2014]. In contrast, the families *Flavobacteriaceae* and *Rhodothermaceae* showed seasonal patterns, with *Flavobacteriaceae* being more pronounced from November to June and *Rhodothermaceae* from June to October (\autoref{bactero}). Within *Alphaproteobacteria* the family *Rhodobacteraceae* comprised the majority of sequences throughout the year (\autoref{alpha}). This metabolically versatile family is often associated with macrophyte surfaces and usually is one of the most abundant groups [@Burke2011a; @Michelou2013; @Pujalte2014; @Manucos2016]. In addition, *Hyphomonadaceae* were found in all samples. Interestingly, some of the species within this group contain stalks on their cells, which can be used to attach to the macrophyte surface [@Weidner2000; @Abraham2014]. Within the *Gammaproteobacteria*, sequences without known representatives were the most pronounced group present throughout the year (\autoref{gamma}). In addition, *Granulosicoccus* was also found in almost all samples. *Gammaproteobacteria* are often a major constituent of macrophyte epiphytic communities [@Michelou2013; @Burke2011a; @Crump2018]. Beside these two groups, other less abundant, taxa showed seasonal and host-specific patterns. For example, *C. cylindracea* harbored *Thioploca*, a known sulfur sediment bacteria and *Cellvibrionaceae*, a family with cultured members known as polysaccharide degraders [@Jorgensen1999; @Xie2017]. *Desulfobacterota* were found only associated with *C. cylindracea* and no group within this phylum showed seasonal patterns (\autoref{desulfo}). The presence of this phylum only on *C. cylindracea* is to be expected as part of the epiphytic community is in direct contact with the sediment. The *Desulfobacterota* community was dominated by *Desulfatitalea* and no relative *Desulfocapsaceae*, known sulfate sediment groups [@Higashioka2015; @Kuever2014].

>>>>

Chloroplast sequence abundances were higher in autumn/winter than in spring/summer. This pattern is not surprising as seagrasses harbor more algal epiphytes during autumn/winter than in spring/summer [@Reyes2001]. Furthermore, we used an adapted DNA isolation protocol that is known to partially co-extract DNA from planktonic eukaryotes [@Korlevic2015]. Strong seasonal fluctuations of high rank epiphytic taxa were not observed, with the exception of *Cyanobacteria*. Cyanobacterial sequences were more pronounced in November and December than in spring and summer. In the months of high cyanobacterial sequence abundances the majority of sequences from this group were classified as *Pleurocapsa*, a group known to colonized different living and non-living surfaces [@Longford2007; @Burns2004; @Mobberley2012; @Reisser2014]. It is possible that during periods of low metabolic activity there is a reduced interaction and selection of the epiphytic community by the seagrass, causing leaves to become a suitable surface for nonspecific colonizers [@Zavodnik1998]. *Pleurocapsa* was replaced in spring and summer by *Acrophormium*, *Phormidesmis* and sequences without known relatives within the *Nodosilineaceae*. A study of coastal microbial mats found also a higher proportion of *Nodosilineaceae* sequences in summer, while *Phormidesmis* sequences were at their peak in autumn [@Cardoso2019]. Other high rank taxa did not exhibit strong successional patterns. In every analyzed group, with the exception of *Desulfobacterota*, taxa present throughout the year in similar proportions and season specific taxa could be identified (Figs. \ref{bactero} and \ref{desulfo}). Within *Bacteroidota* different groups within the family *Saprospiraceae* (e.g. *Lewinella*, *Portibacter* and *Rubidimonas*) were detected across all seasons. Members of this family are often found in association with macrophytes and it is suggested that they are involved in the hydrolysis and utilization of complex carbon sources [@Burke2011a; @Crump2018; @McIlroy2014]. In contrast, the families *Flavobacteriaceae* and *Rhodothermaceae* showed seasonal patterns, with *Flavobacteriaceae* being more pronounced from November to June and *Rhodothermaceae* from June to October (\autoref{bactero}). Within *Alphaproteobacteria* the family *Rhodobacteraceae* comprised the majority of sequences throughout the year (\autoref{alpha}). This metabolically versatile family is often associated with macrophyte surfaces and usually is one of the most abundant groups [@Burke2011a; @Michelou2013; @Pujalte2014; @Mancuso2016]. In addition, *Hyphomonadaceae* were found in all samples. Interestingly, some of the species within this group contain stalks on their cells, which can be used to attach to the macrophyte surface [@Weidner2000; @Abraham2014]. Within the *Gammaproteobacteria*, sequences without known representatives were the most pronounced group present throughout the year (\autoref{gamma}). In addition, *Granulosicoccus* was also found in almost all samples. *Gammaproteobacteria* are often a major constituent of macrophyte epiphytic communities [@Michelou2013; @Burke2011a; @Crump2018]. Beside these two groups, other less abundant, taxa showed seasonal and host-specific patterns. For example, *C. cylindracea* harbored *Thioploca*, a known sulfur sediment bacteria and *Cellvibrionaceae*, a family with cultured members known as polysaccharide degraders [@Jorgensen1999; @Xie2017]. *Desulfobacterota* were found only associated with *C. cylindracea* and no group within this phylum showed seasonal patterns (\autoref{desulfo}). The presence of this phylum only on *C. cylindracea* is to be expected as part of the epiphytic community is in direct contact with the sediment. The *Desulfobacterota* community was dominated by *Desulfatitalea* and no relative *Desulfocapsaceae*, known sulfate sediment groups [@Higashioka2015; @Kuever2014].

### 2020-05-29 submission/manuscript.tex
Review optimization.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,135
different OTUs with 373 to 2,025 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,694.2 ±
133.7 OTUs{]} and monospecific {[}Number of OTUs, 1,737.9 ± 160.5
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,055.3 ± 209.2
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 532.6 ± 141.9 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.4 -- 51.9 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 51.9
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

surfaces and usually is one of the most abundant groups ({\textbf{???}};
Burke and Thomas \emph{et al.}, 2011; Michelou \emph{et al.}, 2013;
Pujalte \emph{et al.}, 2014). In addition, \emph{Hyphomonadaceae} were
found in all samples. Interestingly, some of the species within this
group contain stalks on their cells, which can be used to attach to the
macrophyte surface (Weidner \emph{et al.}, 2000; Abraham and Rohde,
2014). Within the \emph{Gammaproteobacteria}, sequences without known
representatives were the most pronounced group present throughout the
year (\autoref{gamma}). In addition, \emph{Granulosicoccus} was also
found in almost all samples. \emph{Gammaproteobacteria} are often a
major constituent of macrophyte epiphytic communities (Burke and Thomas
\emph{et al.}, 2011; Michelou \emph{et al.}, 2013; Crump \emph{et al.},
2018). Beside these two groups, other less abundant, taxa showed
seasonal and host-specific patterns. For example, \emph{C. cylindracea}
harbored \emph{Thioploca}, a known sulfur sediment bacteria and
\emph{Cellvibrionaceae}, a family with cultured members known as
polysaccharide degraders (Jørgensen and Gallardo, 1999; Xie \emph{et
al.}, 2017). \emph{Desulfobacterota} were found only associated with
\emph{C. cylindracea} and no group within this phylum showed seasonal
patterns (\autoref{desulfo}). The presence of this phylum only on
\emph{C. cylindracea} is to be expected as part of the epiphytic
community is in direct contact with the sediment. The

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,036
different OTUs with 362 to 2,029 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,684.5 ±
131.3 OTUs{]} and monospecific {[}Number of OTUs, 1,727.6 ± 161.5
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,060.2 ± 212.3
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 524.5 ± 137.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.1 -- 52.4 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.53, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 52.4
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

surfaces and usually is one of the most abundant groups (Burke and
Thomas \emph{et al.}, 2011; Michelou \emph{et al.}, 2013; Pujalte
\emph{et al.}, 2014; Mancuso \emph{et al.}, 2016). In addition,
\emph{Hyphomonadaceae} were found in all samples. Interestingly, some of
the species within this group contain stalks on their cells, which can
be used to attach to the macrophyte surface (Weidner \emph{et al.},
2000; Abraham and Rohde, 2014). Within the \emph{Gammaproteobacteria},
sequences without known representatives were the most pronounced group
present throughout the year (\autoref{gamma}). In addition,
\emph{Granulosicoccus} was also found in almost all samples.
\emph{Gammaproteobacteria} are often a major constituent of macrophyte
epiphytic communities (Burke and Thomas \emph{et al.}, 2011; Michelou
\emph{et al.}, 2013; Crump \emph{et al.}, 2018). Beside these two
groups, other less abundant, taxa showed seasonal and host-specific
patterns. For example, \emph{C. cylindracea} harbored \emph{Thioploca},
a known sulfur sediment bacteria and \emph{Cellvibrionaceae}, a family
with cultured members known as polysaccharide degraders (Jørgensen and
Gallardo, 1999; Xie \emph{et al.}, 2017). \emph{Desulfobacterota} were
found only associated with \emph{C. cylindracea} and no group within
this phylum showed seasonal patterns (\autoref{desulfo}). The presence
of this phylum only on \emph{C. cylindracea} is to be expected as part
of the epiphytic community is in direct contact with the sediment. The
